---
title: "bNMF thyroid cancer"
author: "Yon Ho Jee"
date: "2024-07-15"
output: html_document
---
Pipeline - This is a modified pipeline available in https://github.com/gwas-partitioning/bnmf-clustering/blob/master/scripts/bNMF_example_pipeline.R. 
1. Load constructed Z score matrix (Construct_Z_matrix.Rmd)
2. Prepare the final z-score matrix (https://github.com/gwas-partitioning/bnmf-clustering/blob/master/scripts/prep_bNMF.R)
3. Run bNMF and export the results in html (https://github.com/gwas-partitioning/bnmf-clustering/blob/master/scripts/run_bNMF.R)
4. Summarize bNMF output into a table

```{r}
library(data.table);library(dplyr);library(tidyr)
library(ggplot2);library(tidyverse)
```

```{r}
# load requires packages
# install.packages("pacman")
pacman::p_load(tidyverse, data.table, readxl, magrittr, dplyr, strex,
               rstudioapi, DT, kableExtra, GenomicRanges)

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# load project scripts containing bNMF functions - downloaded from https://github.com/gwas-partitioning/bnmf-clustering/tree/master/scripts
source("/bnmf-clustering/scripts/prep_bNMF.R")  # prep_z_matrix
source("/bnmf-clustering/scripts/run_bNMF.R")  # run_bNMF & summarize_bNMF
```

# 1.Load constructed Z score matrix
```{r}
GWAS_sumstats_analyzed <- read_excel("/Thyroid/bNMF/GWAS_sumstats_analyzed.xlsx")

meta_df <- GWAS_sumstats_analyzed %>%
  filter(str_starts(Biobanks, "KCPS2|ThyroidOmics"))
finngen_bbj_df <- GWAS_sumstats_analyzed %>%
  filter(str_starts(Biobanks, "finngen|bbj")) 
ukb_df <- GWAS_sumstats_analyzed %>%
  filter(str_starts(Biobanks, "ukb"))
gwascat_df <- GWAS_sumstats_analyzed %>%
  filter(str_starts(Biobanks, "gwascat"))

trait_list1 <- meta_df$`File name`
n_list1 <- meta_df$N
term_list1 <- meta_df$Phenotype

trait_list2 <- finngen_bbj_df$`File name`
n_list2 <- finngen_bbj_df$N
term_list2 <- finngen_bbj_df$Phenotype

trait_list3 <- str_remove(ukb_df$`File name`, "\\.tsv\\.bgz$")
n_list3 <- ukb_df$N
term_list3 <- ukb_df$Phenotype

trait_list4 <- gwascat_df$`File name`
n_list4 <- gwascat_df$N
term_list4 <- gwascat_df$Phenotype
```

```{r}
sumstat1_leadSNPs <- fread(paste0("/leadSNPs.txt"), data.table = F) %>% #full data available in Supplementary Table 2
  mutate(
    min_allele = pmin(ref, alt),
    max_allele = pmax(ref, alt),
    harmonized_SNP_hg38 = paste(chrom, pos, min_allele, max_allele, sep=":")) %>%
  dplyr::rename(chr=chrom) 

alignment_GWAS_summStats <- sumstat1_leadSNPs %>%
  dplyr::rename(RSID=rsid) %>%
  mutate(SNP=paste0(CHR,":",pos_hg19),
         Risk_Allele = ifelse(BETA>=0, EFFECT_ALLELE, OTHER_ALLELE),
         Nonrisk_Allele = ifelse(BETA>=0, OTHER_ALLELE, EFFECT_ALLELE),
         Risk_Allele_Orig = Risk_Allele,
         ALT=EFFECT_ALLELE, 
         REF=OTHER_ALLELE,
         BETA_new = ifelse(BETA>=0, BETA, -BETA),
         FREQ_new = ifelse(BETA>=0, Freq1, 1-Freq1)) %>%
  dplyr::rename(P_VALUE=P) %>%
  dplyr::select(SNP,Risk_Allele_Orig,ALT,REF,Risk_Allele,Nonrisk_Allele,P_VALUE,BETA_new) %>%
  dplyr::rename(BETA=BETA_new)  
  
#write_csv(x = alignment_GWAS_summStats,
#          file = file.path("/bNMF/test_results/alignment_GWAS_summStats.csv"))
#alignment_GWAS_summStats <- fread("/bNMF/test_results/alignment_GWAS_summStats.csv", data.table = F)
```

```{r}
trait_list <- c(term_list1[-1],term_list2,term_list3,term_list4)

z_mat <- fread(paste0('/tmp_results/z_mat_META_FinnGen_UKB_MVP_GHI_LAT_PMBB_autoreport.txt'), data.table = F)

z_mat <- as.matrix(z_mat[,-1])
rownames(z_mat) <- alignment_GWAS_summStats$SNP
colnames(z_mat) <- trait_list

N_pheno <- fread(paste0('/tmp_results/N_pheno_META_FinnGen_UKB_MVP_GHI_LAT_PMBB_autoreport.txt'), data.table = F)

N_mat <- as.matrix(N_pheno, nrow=nloci, ncol=ncol(N_pheno))
N_mat <- as.matrix(N_mat[,-1])
rownames(N_mat) <- alignment_GWAS_summStats$SNP
colnames(N_mat) <- trait_list

initial_zscore_matrices_final <- list(z_mat=z_mat,N_mat=N_mat)
```

# 2. Prepare the final z-score matrix (Section 10. Generate non-negative z-score matrix)
```{r}
prep_z_output <- prep_z_matrix(z_mat = initial_zscore_matrices_final$z_mat,
                               N_mat = initial_zscore_matrices_final$N_mat,
                               corr_cutoff = 0.8)

#   1.) The scaled, non-negative z-score matrix
final_zscore_matrix <- prep_z_output$final_z_mat

#   2.) Results from the trait filtering
df_traits_filtered <- prep_z_output$df_traits
write_csv(x = df_traits_filtered,
          file = file.path(project_dir,"df_traits.csv"))

# prep_z_matrix also save trait correlation matrix to working dir, so move to project dir
system(sprintf("mv trait_cor_mat.txt %s", project_dir))

print(sprintf("Final matrix: %i SNPs x %i traits",
      nrow(final_zscore_matrix),
      ncol(final_zscore_matrix)/2))

save.image(file = file.path(project_dir, "pipeline_data.RData"))
```

# 3. Run bNMF and summarize the results (Section 11. Run bNMF) 
```{r}
start=Sys.time()

bnmf_reps <- run_bNMF(final_zscore_matrix,
                      n_reps=100,
                      tolerance = 1e-6)
summarize_bNMF(bnmf_reps, dir_save=project_dir)

save.image(file = file.path(project_dir, "pipeline_data.RData"))

end=Sys.time()
print("Total pipeline runtime:")
print(end-start)
#----
```
## Convergence plots for bNMF runs using n.error and n.lambda
```{r}
library(tidyverse)

summarize_bnmf_reps <- function(bnmf_reps) {
  map_dfr(seq_along(bnmf_reps), function(r) {
    res <- bnmf_reps[[r]]
    
    final_iter <- length(res$n.lambda)
    final_lambda <- res$n.lambda[[final_iter]]
    
    tibble(
      run = r,
      iterations = final_iter,
      final_error = res$n.error[[final_iter]],
      final_evidence = res$n.evid[[final_iter]],
      K = sum(final_lambda > min(final_lambda)),  # effective K
      lambda_min = min(final_lambda),
      lambda_median = median(final_lambda),
      lambda_max = max(final_lambda)
    )
  }) %>% arrange(final_evidence)
}

run_summary <- summarize_bnmf_reps(bnmf_reps)
head(run_summary)
```
```{r}
library(tidyverse)

bnmf_run_to_df_safe <- function(res, run_id = 1,
                               baseline = c("iter_min", "final_min", "none"),
                               cost_source = c("n.evid", "n.error")) {
  baseline <- match.arg(baseline)
  cost_source <- match.arg(cost_source)

  # ---- lambdas ----
  lam_list <- res$n.lambda
  lam_list <- lam_list[!sapply(lam_list, is.null)]
  lam_mat <- do.call(rbind, lapply(lam_list, as.numeric))
  it_lam <- seq_len(nrow(lam_mat))
  K <- ncol(lam_mat)
  colnames(lam_mat) <- paste0("k", seq_len(K))

  # baseline B for lambda_k - B
  if (baseline == "iter_min") {
    B_vec <- apply(lam_mat, 1, min, na.rm = TRUE)
  } else if (baseline == "final_min") {
    B_final <- min(lam_mat[nrow(lam_mat), ], na.rm = TRUE)
    B_vec <- rep(B_final, nrow(lam_mat))
  } else {
    B_vec <- rep(0, nrow(lam_mat))
  }

  df_lambda <- as_tibble(lam_mat) %>%
    mutate(iter = it_lam, run = run_id, B = B_vec) %>%
    pivot_longer(cols = starts_with("k"),
                 names_to = "component",
                 values_to = "lambda") %>%
    mutate(lambda_minus_B = lambda - B)

  # ---- cost ----
  cost_list <- res[[cost_source]]
  cost_list <- cost_list[!sapply(cost_list, is.null)]
  cost_vec <- unlist(cost_list)

  df_cost <- tibble(
    run = run_id,
    iter = seq_along(cost_vec),
    cost = cost_vec
  )

  list(df_cost = df_cost, df_lambda = df_lambda)
}

plot_convergence <- function(bnmf_reps,
                                       run_id = 1,
                                       baseline = c("iter_min", "final_min", "none"),
                                       out_prefix = paste0("bnmf_run", run_id),
                                       use_lambda_minus_B = TRUE,
                                       cost_source = c("n.evid", "n.error")) {
  baseline <- match.arg(baseline)
  cost_source <- match.arg(cost_source)

  res <- bnmf_reps[[run_id]]
  dfs <- bnmf_run_to_df_safe(res, run_id = run_id, baseline = baseline, cost_source = cost_source)
  df_cost <- dfs$df_cost
  df_lam  <- dfs$df_lambda

  # Middle: cost (log-log)
  p_cost <- ggplot(df_cost, aes(x = iter, y = cost)) +
    geom_line(linewidth = 0.6) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
      title = paste0("cost (run ", run_id, ")"),
      x = "",
      y = paste0(cost_source, " (log-log)")
    ) +
    theme_bw()

  # Bottom: relevance (log-linear)
  yvar <- if (use_lambda_minus_B) "lambda_minus_B" else "lambda"
  ylab <- if (use_lambda_minus_B) expression(lambda[k] - B) else expression(lambda[k])

  p_rel <- ggplot(df_lam, aes(x = iter, y = .data[[yvar]], group = component)) +
    geom_line(alpha = 0.7, linewidth = 0.35) +
    scale_x_log10() +
    labs(
      title = paste0("relevance (run ", run_id, ")"),
      x = "iterations",
      y = ylab
    ) +
    theme_bw()

  list(cost_plot = p_cost, relevance_plot = p_rel)
}

library(cowplot)

plots <- plot_convergence(
  bnmf_reps,
  run_id = 1,
  baseline = "iter_min",
  out_prefix = "bnmf_run1",
  use_lambda_minus_B = TRUE,
  cost_source = "n.evid"
)

p_stacked <- plot_grid(
  plots$cost_plot,
  plots$relevance_plot,
  ncol = 1,
  align = "v",
  rel_heights = c(1, 1.1)
)

ggsave(
  "~/bNMF/bnmf_run1_convergence_stacked.pdf",
  p_stacked,
  width = 7,
  height = 6
)

p_stacked

#check if bNMF converges well before 10,000 iterations across runs
ggplot(run_summary, aes(x = run, y = iterations)) +
   geom_point(size = 2) + geom_hline(yintercept = 10000, linetype = "dashed", color = "red") +
   labs(x = "Run (random initialization)", y = "Iterations to convergence") +
   theme_bw()
```

```{r}
gwas_traits <- NULL

meta_studies = "KCPS2_GBMI_FinnGen_MVP_ukb_out_GHI_LAT_PMBB"

rsID_map <- sumstat1_leadSNPs %>%
  left_join(meta_res,by=c("harmonized_SNP_hg38"="SNP")) %>%
  left_join(sumstat1_leadSNPs_hg19 %>% dplyr::select(harmonized_SNP_hg38, harmonized_SNP,pos_hg19), by ="harmonized_SNP_hg38") %>%
  dplyr::rename(rsID=rsid) %>%
  mutate(VAR_ID=paste0(CHR,"_",pos_hg19,"_",OTHER_ALLELE,"_",EFFECT_ALLELE)) %>%
  dplyr::select(VAR_ID,rsID)
fwrite(rsID_map,file.path(project_dir, "rsID_map.txt"),sep = "\t")
```

```{r}
setwd("/bNMF")

# format results
k <- NULL; date <- today() 
if (is.null(k)){
  html_filename <- sprintf("results_for_maxK_%s.html", date)
} else {
  html_filename <- sprintf("results_for_K_%i_%s.html", k, date)
}
# https://github.com/gwas-partitioning/bnmf-clustering/blob/master/scripts/format_bNMF_results.Rmd
rmarkdown::render(
  '/bnmf-clustering/scripts/format_bNMF_results.Rmd',
  output_file = html_filename,
  params = list(main_dir = project_dir,
                k = k,
                loci_file="query",
                GTEx=F,
                my_traits=gwas_traits)
)
```
/bnmf-clustering/scripts/results_for_maxK.html
/bNMF/test_results

# 4. Summarize bNMF output into a table
```{r}
out_df <- read_excel(paste0(project_dir,"/sorted_cluster_weights_K",k,"_rev.xlsx")) 
var_w <- out_df %>%
  filter(!is.na(VAR_ID)) %>%
  dplyr::select(1:(k+3)) %>%
  dplyr::rename_at(vars(4:(k+3)), ~paste0("Cluster", 1:k))
pheno_w <- out_df %>%
  filter(!is.na(trait)) %>%
  dplyr::select((k+6):(k+6+k)) %>%
 mutate(trait = gsub("_pos$", "(+)", trait),
        trait = gsub("_neg$", "(-)", trait)) %>%
  dplyr::rename_at(vars(2:(k+1)), ~paste0("Cluster", 1:k))
```

```{r}
#
# Initialize an empty list to store the results
cluster_summary <- lapply(1:k, function(cluster) {
  cluster_col <- paste0("Cluster", cluster)
  
  # Number of variants for the cluster
  num_variants <- sum(var_w[[cluster_col]] > cutoff, na.rm = TRUE)
  
  # Key top-weighted traits for the cluster
  top_traits <- pheno_w %>%
    filter(!!sym(cluster_col) > cutoff) %>%
    arrange(desc(!!sym(cluster_col))) %>%
    pull(trait)
  
  # Key top-weighted loci for the cluster
  top_loci <- var_w %>%
    filter(!!sym(cluster_col) > cutoff) %>%
    arrange(desc(!!sym(cluster_col))) %>%
    pull(gene) %>%
    unique()
  
  # Return the results for the cluster
  data.frame(
    Cluster = cluster_col,
    `Number of variants` = num_variants,
    `Key top-weighted traits` = paste(top_traits, collapse = ", "),
    `Key top-weighted loci` = paste(top_loci, collapse = ", ")
  )
})

# Combine results for all clusters into a single data frame
final_table <- do.call(rbind, cluster_summary)

# Optionally, save the table to a CSV or TSV file
write.csv(final_table, paste0(project_dir,"/cluster_summary_table_",k,"_",date,".csv"), row.names = FALSE)
```
